How important are functional and developmental constraints on phenotypic evolution? An empirical test with the stomatal anatomy of flowering plants

### 2023-09-06 ms/ms.Rmd
final tweaks after publication

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied [@bergmann_stomatal_2007; @dow_patterning_2014; @sack_developmental_2016]. We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes S1. A graphical summary is provided in Fig. \@ref(fig:developmental-integration). We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division [@dow_patterning_2014]. The second parameter is stomatal index [@salisbury_i_1928; @sack_developmental_2016], which is determined by amplifying and spacing divisions after asymmetric cell division [@dow_patterning_2014]. When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density [see @lynch_genetics_1998 for key random variable algebra theorems].

@@

We generated an ultrametric, bifurcating phylogeny of `r n_tax3` taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014], and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration [see @smith_constructing_2018; @magallon_metacalibrated_2015 for further detail]. The complete methodology is described in the online supplement (Notes S2).

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes S1 and subsections below). Within and among species covariation is a hallmark of developmental integration [@armbruster_multilevel_1988], but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the $4 \times 4$ covariance matrix of phylogenetically independent contrasts between log-transformed values of $\Delta \text{log}(D_\mathrm{ab})$, $\Delta \text{log}(D_\mathrm{ad})$, $\Delta \text{log}(L_\mathrm{ab})$, and $\Delta \text{log}(L_\mathrm{ad})$ using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as $\text{Var}[\Delta \text{log(trait)}]$ and covariances as $\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]$. We used a multivariate $t$-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values [@lange_robust_1989]. We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g. Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see Notes S1 for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See Notes S1 and Table \@ref(tab:predictions) for more a complete derivation and detailed predictions.

@@

Functional and developmental constraints may slow adaptation by preventing traits from evolving independently towards a multivariate phenotypic optimum. An alternative view is that phenotypic evolution is limited primarily because most phenotypes are maladaptive and selected against, sometimes referred to as selective constraint. In this study, we took advantage of the fact that amphistomatous leaves produce stomata on both abaxial (usually lower) and adaxial (usually upper) surfaces. Packing limits, a functional constraint, and developmental integration should result in similar (co)variance in divergence of stomatal traits on each surface (Notes S1), whereas differing selective pressures for each surface would lead to different patterns of divergence. Although some patterns of trait (co)variance are compatible with packing limits or developmental integration (see below), independent evolution of stomatal density on each surface across flowering plants is notably inconsistent with these constraints. We therefore conclude that selection against generally maladaptive trait combinations plays a major role in the evolution of stomata and possibly other ecologically important traits. It should not be surprising that stomatal traits respond to natural selection, but it is nevertheless plausible that some stomatal trait combinations that would have high fitness are precluded by nonadaptive constraints. Our results challenge these nonadaptive explanations, especially for stomatal density.

@@

If meristematic cell volume and expansion integrate stomatal size and density [@brodribb_unified_2013], then we predicted inverse size-density scaling would evolve with the same (co)variance for both ab- and adaxial leaf surfaces (Notes S1). Contrary to this prediction, there are many combinations of stomatal density ($D_i$) and length ($L_i$) found on adaxial leaf surfaces that are absent from abaxial leaf surfaces (Fig. \ref{fig:h1_raw}). In principle, the different relationship between traits on each surface could be caused by different evolutionary variance in stomatal density ($\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})]$) and/or covariance ($\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]$) on each surface. However, the covariance relationship between density and length is similar on each surface, whereas the evolutionary variance in adaxial stomatal density is significantly higher than that for abaxial density ($\text{Var}[\Delta \text{log}(D_\text{ab})] < \text{Var}[\Delta \text{log}(D_\text{ad})]$; Fig. \ref{fig:h1}). Given that the average stomatal length is usually about the same on each surface (see below), these results imply that plants can often evolve stomatal densities on each surface without a concomitant change in size. Based on our theoretical analysis, we interpret these results to mean that cell divisions affecting stomatal index are less evolutionarily constrained than the asymmetric cell division preceding the guard cell meristemoid (Fig. \ref{fig:developmental-integration}; Table \ref{tab:predictions})

>>>>

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied [@bergmann_stomatal_2007; @dow_patterning_2014; @sack_developmental_2016]. We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to the online supplement. A graphical summary is provided in Fig. \@ref(fig:developmental-integration). We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division [@dow_patterning_2014]. The second parameter is stomatal index [@salisbury_i_1928; @sack_developmental_2016], which is determined by amplifying and spacing divisions after asymmetric cell division [@dow_patterning_2014]. When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density [see @lynch_genetics_1998 for key random variable algebra theorems].

@@

We generated an ultrametric, bifurcating phylogeny of `r n_tax3` taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants [@smith_constructing_2018; @zanne_three_2014], and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration [see @smith_constructing_2018; @magallon_metacalibrated_2015 for further detail]. The complete methodology is described in the online supplement.

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see the online supplement and subsections below). Within and among species covariation is a hallmark of developmental integration [@armbruster_multilevel_1988], but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the $4 \times 4$ covariance matrix of phylogenetically independent contrasts between log-transformed values of $\Delta \text{log}(D_\mathrm{ab})$, $\Delta \text{log}(D_\mathrm{ad})$, $\Delta \text{log}(L_\mathrm{ab})$, and $\Delta \text{log}(L_\mathrm{ad})$ using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as $\text{Var}[\Delta \text{log(trait)}]$ and covariances as $\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]$. We used a multivariate $t$-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values [@lange_robust_1989]. We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g. Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see the online supplement for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See the online supplement and Table \@ref(tab:predictions) for more a complete derivation and detailed predictions.

@@

Functional and developmental constraints may slow adaptation by preventing traits from evolving independently towards a multivariate phenotypic optimum. An alternative view is that phenotypic evolution is limited primarily because most phenotypes are maladaptive and selected against, sometimes referred to as selective constraint. In this study, we took advantage of the fact that amphistomatous leaves produce stomata on both abaxial (usually lower) and adaxial (usually upper) surfaces. Packing limits, a functional constraint, and developmental integration should result in similar (co)variance in divergence of stomatal traits on each surface (see the online supplement), whereas differing selective pressures for each surface would lead to different patterns of divergence. Although some patterns of trait (co)variance are compatible with packing limits or developmental integration (see below), independent evolution of stomatal density on each surface across flowering plants is notably inconsistent with these constraints. We therefore conclude that selection against generally maladaptive trait combinations plays a major role in the evolution of stomata and possibly other ecologically important traits. It should not be surprising that stomatal traits respond to natural selection, but it is nevertheless plausible that some stomatal trait combinations that would have high fitness are precluded by nonadaptive constraints. Our results challenge these nonadaptive explanations, especially for stomatal density.

@@

If meristematic cell volume and expansion integrate stomatal size and density [@brodribb_unified_2013], then we predicted inverse size-density scaling would evolve with the same (co)variance for both ab- and adaxial leaf surfaces (see the online supplement). Contrary to this prediction, there are many combinations of stomatal density ($D_i$) and length ($L_i$) found on adaxial leaf surfaces that are absent from abaxial leaf surfaces (Fig. \ref{fig:h1_raw}). In principle, the different relationship between traits on each surface could be caused by different evolutionary variance in stomatal density ($\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})]$) and/or covariance ($\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]$) on each surface. However, the covariance relationship between density and length is similar on each surface, whereas the evolutionary variance in adaxial stomatal density is significantly higher than that for abaxial density ($\text{Var}[\Delta \text{log}(D_\text{ab})] < \text{Var}[\Delta \text{log}(D_\text{ad})]$; Fig. \ref{fig:h1}). Given that the average stomatal length is usually about the same on each surface (see below), these results imply that plants can often evolve stomatal densities on each surface without a concomitant change in size. Based on our theoretical analysis, we interpret these results to mean that cell divisions affecting stomatal index are less evolutionarily constrained than the asymmetric cell division preceding the guard cell meristemoid (Fig. \ref{fig:developmental-integration}; Table \ref{tab:predictions})

### 2023-09-06 ms/ms.tex
final tweaks after publication

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in \emph{R} version 4.2.1 (\protect\hyperlink{ref-r_core_team_r:_2022}{R Core Team 2022}). Source code is publicly available on GitHub (\url{https://github.com/cdmuir/stomata-independence}) and archived on Zenodo (see Data availability).

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied (\protect\hyperlink{ref-bergmann_stomatal_2007}{Bergmann and Sack 2007}; \protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to Notes S1. A graphical summary is provided in Fig. \ref{fig:developmental-integration}. We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). The second parameter is stomatal index (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}), which is determined by amplifying and spacing divisions after asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density (see \protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh 1998} for key random variable algebra theorems).

@@

We generated an ultrametric, bifurcating phylogeny of 638 taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}), and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration (see \protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-magallon_metacalibrated_2015}{Magallón et al. 2015} for further detail). The complete methodology is described in the online supplement (Notes S2).

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see Notes S1 and subsections below). Within and among species covariation is a hallmark of developmental integration (\protect\hyperlink{ref-armbruster_multilevel_1988}{Armbruster 1988}), but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the \(4 \times 4\) covariance matrix of phylogenetically independent contrasts between log-transformed values of \(\Delta \text{log}(D_\mathrm{ab})\), \(\Delta \text{log}(D_\mathrm{ad})\), \(\Delta \text{log}(L_\mathrm{ab})\), and \(\Delta \text{log}(L_\mathrm{ad})\) using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as \(\text{Var}[\Delta \text{log(trait)}]\) and covariances as \(\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]\). We used a multivariate \(t\)-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values (\protect\hyperlink{ref-lange_robust_1989}{Lange, Little, and Taylor 1989}). We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g.~Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see Notes S1 for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See Notes S1 and Table \ref{tab:predictions} for more a complete derivation and detailed predictions.

@@

If meristematic cell volume and expansion integrate stomatal size and density (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}), then we predicted inverse size-density scaling would evolve with the same (co)variance for both ab- and adaxial leaf surfaces (Notes S1). Contrary to this prediction, there are many combinations of stomatal density (\(D_i\)) and length (\(L_i\)) found on adaxial leaf surfaces that are absent from abaxial leaf surfaces (Fig. \ref{fig:h1_raw}). In principle, the different relationship between traits on each surface could be caused by different evolutionary variance in stomatal density (\(\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})]\)) and/or covariance (\(\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]\)) on each surface. However, the covariance relationship between density and length is similar on each surface, whereas the evolutionary variance in adaxial stomatal density is significantly higher than that for abaxial density (\(\text{Var}[\Delta \text{log}(D_\text{ab})] < \text{Var}[\Delta \text{log}(D_\text{ad})]\); Fig. \ref{fig:h1}). Given that the average stomatal length is usually about the same on each surface (see below), these results imply that plants can often evolve stomatal densities on each surface without a concomitant change in size. Based on our theoretical analysis, we interpret these results to mean that cell divisions affecting stomatal index are less evolutionarily constrained than the asymmetric cell division preceding the guard cell meristemoid (Fig. \ref{fig:developmental-integration}; Table \ref{tab:predictions})

@@

\leavevmode\vadjust pre{\hypertarget{ref-haworth_co-ordination_2013}{}}%
Haworth, Matthew, Caroline Elliott-Kingston, and Jennifer C. McElwain. 2013. {``Co-Ordination of Physiological and Morphological Responses of Stomata to Elevated {[}{Co2}{]} in Vascular Plants.''} \emph{Oecologia} 171 (1): 71--82. \url{https://doi.org/10.1007/s00442-012-2406-9}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lake_longdistance_2002}{}}%
Lake, Janice A., F. Ian Woodward, and W. Paul Quick. 2002. {``Long‐distance {Co2} Signalling in Plants.''} \emph{Journal of Experimental Botany} 53 (367): 183--93. \url{https://doi.org/10.1093/jexbot/53.367.183}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rae_elucidating_2006}{}}%
Rae, A. M., Rachel Ferris, M. J. Tallis, and Gail Taylor. 2006. {``Elucidating Genomic Regions Determining Enhanced Leaf Growth and Delayed Senescence in Elevated {Co2}.''} \emph{Plant, Cell and Environment} 29 (9): 1730--41. \url{https://doi.org/10.1111/j.1365-3040.2006.01545.x}.

>>>>

Unless otherwise mentioned, we performed all data wrangling and statistical analyses in \emph{R} version 4.2.2 (\protect\hyperlink{ref-r_core_team_r:_2022}{R Core Team 2022}). Source code is publicly available on GitHub (\url{https://github.com/cdmuir/stomata-independence}) and archived on Zenodo (see Data availability).

@@

Developmental integration could shape patterns of phenotypic macroevolution, but a major hindrance to progress is that verbal models do not make precise, quantitative predictions that distinguish it from alternatives. An advantage of testing developmental integration in stomata is that their development is well studied (\protect\hyperlink{ref-bergmann_stomatal_2007}{Bergmann and Sack 2007}; \protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}). We can leverage that knowledge to build a developmental function and derive equations for phenotypic (co)variance caused by developmental integration. If observed patterns of evolution are inconsistent with developmental integration, theory may also help identify which parameters lead to developmental disintegration. We combined and extended stomatal development models to predict how stomatal density and length would diverge if stomatal development were constrained and how those predictions would change if stomatal development were unconstrained. We summarize our methods verbally here and direct readers interested in the mathematical details to the online supplement. A graphical summary is provided in Fig. \ref{fig:developmental-integration}. We imposed constraint by assuming the stomatal developmental function is constrained. The developmental function maps cell size prior to differentiation onto stomatal size and density using two parameters. The first parameter describes how cell volume is apportioned between epidermal cells and guard cell meristemoids during asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). The second parameter is stomatal index (\protect\hyperlink{ref-salisbury_i_1928}{Salisbury 1928}; \protect\hyperlink{ref-sack_developmental_2016}{Sack and Buckley 2016}), which is determined by amplifying and spacing divisions after asymmetric cell division (\protect\hyperlink{ref-dow_patterning_2014}{Dow and Bergmann 2014}). When these parameters are fixed, divergence in stomatal size and density is determined by divergence in meristematic cell volume and expansion prior to asymmetric division. We relaxed this constraint by treating parameters of the developmental function as random variables that can diverge between species. We used random variable algebra to derive predicted (co)variance in divergence between stomatal length and density (see \protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh 1998} for key random variable algebra theorems).

@@

We generated an ultrametric, bifurcating phylogeny of 638 taxa by resolving and removing ambiguous taxonomic names, placing taxa on the GBOTB.extended mega-tree of seed plants (\protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-zanne_three_2014}{Zanne et al. 2014}), and resolving polytomies using published sequence data. Divergence times in this phylogeny are based on extensive fossil calibration (see \protect\hyperlink{ref-smith_constructing_2018}{S. A. Smith and Brown 2018}; \protect\hyperlink{ref-magallon_metacalibrated_2015}{Magallón et al. 2015} for further detail). The complete methodology is described in the online supplement.

@@

All hypotheses make predictions about trait (co)variance matrices or parameters derived from them (see the online supplement and subsections below). Within and among species covariation is a hallmark of developmental integration (\protect\hyperlink{ref-armbruster_multilevel_1988}{Armbruster 1988}), but other evolutionary processes also lead to covariance. Distinguishing between them requires deriving predictions and testing whether observed covariance is consistent with one hypothesis or another. We estimated the \(4 \times 4\) covariance matrix of phylogenetically independent contrasts between log-transformed values of \(\Delta \text{log}(D_\mathrm{ab})\), \(\Delta \text{log}(D_\mathrm{ad})\), \(\Delta \text{log}(L_\mathrm{ab})\), and \(\Delta \text{log}(L_\mathrm{ad})\) using a distributional multiresponse robust Bayesian approach. See Table \ref{tab:traits} for variable definitions. We denote variances as \(\text{Var}[\Delta \text{log(trait)}]\) and covariances as \(\text{Cov}[\Delta \text{log(trait}_1),\Delta \text{log(trait}_2)]\). We used a multivariate \(t\)-distribution rather than a Normal distribution because estimates using the former are more robust to exceptional trait values (\protect\hyperlink{ref-lange_robust_1989}{Lange, Little, and Taylor 1989}). We also estimated whether the variance in trait divergence increases with time. Under many trait evolution models (e.g.~Brownian motion), interspecific variance increases through time. To account for this, we included time since taxon-pair divergence as an explanatory variable affecting the trait covariance matrix.

@@

We asked how divergence in stomatal length and density would covary if the developmental function were constrained and compared it to their divergence when the developmental function can evolve. When the developmental function is constrained this means that allocation to guard cell meristemoids during asymmetric division and stomatal index are fixed (see the online supplement for mathematical description). Under these assumptions, divergence in stomatal length and density is mediated entirely by divergence in meristematic cell volume and expansion prior to differentiation. Developmental integration is strong because divergence in density is perfectly negatively correlated with divergence in size. In contrast, stomatal length and density can diverge independently when the developmental function is not fixed. Divergence in asymmetric cell division affects stomatal size independently of density; divergence in stomatal index affects stomatal density independently of size. Divergence in the developmental function causes developmental disintegration because stomatal density and size can diverge independently. Developmental integration is minimal when asymmetric cell division and/or stomatal index diverge more than meristematic cell volume and expansion. The three main conclusions are that 1) developmental constraint leads to developmental integration; 2) different (co)variance in divergence of stomatal length and density on each surface implies the developmental function is not fixed; and 3) divergence in different components of the developmental function affect stomatal length and density differently. See the online supplement and Table \ref{tab:predictions} for more a complete derivation and detailed predictions.

@@

If meristematic cell volume and expansion integrate stomatal size and density (\protect\hyperlink{ref-brodribb_unified_2013}{Brodribb, Jordan, and Carpenter 2013}), then we predicted inverse size-density scaling would evolve with the same (co)variance for both ab- and adaxial leaf surfaces (see the online supplement). Contrary to this prediction, there are many combinations of stomatal density (\(D_i\)) and length (\(L_i\)) found on adaxial leaf surfaces that are absent from abaxial leaf surfaces (Fig. \ref{fig:h1_raw}). In principle, the different relationship between traits on each surface could be caused by different evolutionary variance in stomatal density (\(\text{Var}[\Delta \text{log}(D_\text{ab})] \ne \text{Var}[\Delta \text{log}(D_\text{ad})]\)) and/or covariance (\(\text{Cov}[\Delta \text{log}(L_\text{ab}), \Delta \text{log}(D_\text{ab})] \ne \text{Cov}[\Delta \text{log}(L_\text{ad}), \Delta \text{log}(D_\text{ad})]\)) on each surface. However, the covariance relationship between density and length is similar on each surface, whereas the evolutionary variance in adaxial stomatal density is significantly higher than that for abaxial density (\(\text{Var}[\Delta \text{log}(D_\text{ab})] < \text{Var}[\Delta \text{log}(D_\text{ad})]\); Fig. \ref{fig:h1}). Given that the average stomatal length is usually about the same on each surface (see below), these results imply that plants can often evolve stomatal densities on each surface without a concomitant change in size. Based on our theoretical analysis, we interpret these results to mean that cell divisions affecting stomatal index are less evolutionarily constrained than the asymmetric cell division preceding the guard cell meristemoid (Fig. \ref{fig:developmental-integration}; Table \ref{tab:predictions})

@@

\leavevmode\vadjust pre{\hypertarget{ref-haworth_co-ordination_2013}{}}%
Haworth, Matthew, Caroline Elliott-Kingston, and Jennifer C. McElwain. 2013. {``Co-Ordination of Physiological and Morphological Responses of Stomata to Elevated {[}{CO2}{]} in Vascular Plants.''} \emph{Oecologia} 171 (1): 71--82. \url{https://doi.org/10.1007/s00442-012-2406-9}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lake_longdistance_2002}{}}%
Lake, Janice A., F. Ian Woodward, and W. Paul Quick. 2002. {``Long‐distance {CO2} Signalling in Plants.''} \emph{Journal of Experimental Botany} 53 (367): 183--93. \url{https://doi.org/10.1093/jexbot/53.367.183}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-rae_elucidating_2006}{}}%
Rae, A. M., Rachel Ferris, M. J. Tallis, and Gail Taylor. 2006. {``Elucidating Genomic Regions Determining Enhanced Leaf Growth and Delayed Senescence in Elevated {CO2}.''} \emph{Plant, Cell and Environment} 29 (9): 1730--41. \url{https://doi.org/10.1111/j.1365-3040.2006.01545.x}.
